% AQA Harvard Referencing Style Template
% For English Academic Writing
\documentclass[12pt, a4paper, twoside]{article}

% Essential Packages
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{harvard}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{enumitem}
\usepackage{setspace}

% Page Settings
\geometry{
  left=2.5cm,
  right=2.5cm,
  top=2.5cm,
  bottom=2.5cm,
  headheight=15pt,
  headsep=1cm,
  footskip=1cm
}

% Line Spacing
\onehalfspacing

% Title Settings
\title{\textbf{Paper Title}}
\author{Author Name}
\date{\today}

\begin{document}

% Title Page
\maketitle

% Abstract
\begin{abstract}
  This paper introduces... (Abstract content)
\end{abstract}

% Keywords
\textbf{Keywords:} Keyword 1, Keyword 2, Keyword 3

% Table of Contents
\tableofcontents
\newpage

% Main Body
\section{Introduction}
Classical mechanics is deterministic in form, yet many deterministic systems are not predictably simple in practice. Small differences in initial conditions can produce strongly divergent trajectories in non-linear dynamics, so mathematically well-defined equations may still generate motion that appears irregular over finite observation windows. This contrast between formal structure and physical complexity motivates the present investigation.

The double pendulum is an appropriate system for this purpose. It consists of two point masses connected by two rigid, massless strings, with the upper mass attached to a pivot. In the small-angle limit, its behaviour can be analysed through linear normal-mode theory; outside that limit, non-linear coupling leads to richer responses. In this project, the focus is not chaos as an isolated topic, but the broader non-linear response of a \emph{periodically driven} double pendulum, including transitions between regular and irregular motion as control parameters vary.

The study addresses three linked questions. First, how accurately do small-angle analytical predictions describe measured or simulated frequencies in weak-amplitude motion? Second, how do forcing frequency and forcing amplitude influence the qualitative response of the driven system? Third, to what extent does including experimentally identified damping improve agreement between theoretical and numerical results? These questions establish a direct connection between analytical mechanics, numerical modelling, and empirical parameter estimation.

The methodology integrates theory, simulation, and experiment. A Lagrangian formulation is used to derive the equations of motion, and linearisation is applied to obtain the small-angle modal frequencies as reference quantities. Numerical integration is then performed in Python with a Runge-Kutta scheme, while parameter sweeps are carried out over selected ranges of initial conditions and driving parameters. In parallel, a single-pendulum experiment is used to estimate the local damping coefficient, which is subsequently inserted into the driven double-pendulum model. System response is assessed through time series, phase-space trajectories, and frequency-domain analysis, allowing quantitative validation of the small-angle approximation and critical evaluation of model limitations.

The remainder of this report is organised as follows. Section 2 reviews relevant literature on non-linear oscillators, driven pendulum dynamics, and deterministic sensitivity. Section 3 presents the theoretical framework, including Lagrangian derivation, linear modal analysis, and superposition. Section 4 describes numerical implementation and experimental damping identification. Section 5 reports simulation and validation results, and Section 6 concludes with limitations and future extensions.

\section{Literature Review}
  Literature review content...

\section{Theoretical Analysis}
To thoroughly investigate the dynamics of the double pendulum, a robust mathematical model must be established. This section derives the equations of motion using Lagrangian mechanics, applies small-angle approximations to find analytical solutions, and explores the physical implications of the system's normal modes.

\subsection{Kinematics and the Lagrangian}
Consider a double pendulum consisting of two point masses, $m_1$ and $m_2$, suspended by two massless rigid rods of lengths $l_1$ and $l_2$. Let $\theta_1$ and $\theta_2$ be the angles the rods make with the vertical downward direction. The Cartesian coordinates of the masses, taking the suspension point as the origin and the downward direction as the positive $y$-axis, are given by:
\begin{align*}
x_1 &= l_1 \sin \theta_1, \quad y_1 = l_1 \cos \theta_1, \\
x_2 &= l_1 \sin \theta_1 + l_2 \sin \theta_2, \quad y_2 = l_1 \cos \theta_1 + l_2 \cos \theta_2.
\end{align*}

The kinetic energy ($T$) of the system is the sum of the kinetic energies of both masses, and the potential energy ($V$) is determined by their vertical positions in a uniform gravitational field $g$:
\begin{align*}
T &= \frac{1}{2}m_1(\dot{x}_1^2 + \dot{y}_1^2) + \frac{1}{2}m_2(\dot{x}_2^2 + \dot{y}_2^2), \\
V &= -m_1 g y_1 - m_2 g y_2.
\end{align*}

By substituting the time derivatives of the coordinates into the kinetic energy equation, the Lagrangian of the system, defined as $L = T - V$, is constructed:
\begin{equation}
\begin{aligned}
L &= \frac{1}{2}(m_1 + m_2)l_1^2\dot{\theta}_1^2 + \frac{1}{2}m_2 l_2^2 \dot{\theta}_2^2 + m_2 l_1 l_2 \dot{\theta}_1 \dot{\theta}_2 \cos(\theta_1 - \theta_2) \\
  &\quad + (m_1 + m_2)g l_1 \cos\theta_1 + m_2 g l_2 \cos\theta_2.
\end{aligned}
\end{equation}

\subsection{Non-linear Coupling, Fictitious Forces, and Symplectic Integration}
A critical feature of the Lagrangian derived above is the kinetic energy coupling term, $m_2 l_1 l_2 \dot{\theta}_1 \dot{\theta}_2 \cos(\theta_1 - \theta_2)$. This term indicates that the kinetic energy depends not only on the generalized velocities ($\dot{\theta}_1, \dot{\theta}_2$) but also on the generalized coordinates (specifically the angle difference $\theta_1 - \theta_2$). Consequently, the mass matrix of the system is configuration-dependent, making the system fundamentally non-separable.

When applying the Euler-Lagrange equations, the derivatives of this coupling term naturally give rise to terms that physically represent fictitious forces in the non-inertial rotational reference frame:
\begin{itemize}[leftmargin=*]
    \item \textbf{Centrifugal Force:} Taking the partial derivative of the Lagrangian with respect to the position coordinate, $\frac{\partial L}{\partial \theta_1}$, produces terms proportional to the square of the angular velocities (e.g., $\dot{\theta}_2^2$). Physically, this represents the centrifugal force, an outward inertial pull exerted by the rotation of one pendulum upon the other.
    \item \textbf{Coriolis Force:} Taking the time derivative of the generalized momentum, $\frac{d}{dt}\left(\frac{\partial L}{\partial \dot{\theta}_1}\right)$, requires applying the chain rule to the $\cos(\theta_1 - \theta_2)$ term. This generates cross-product velocity terms (e.g., $\dot{\theta}_1 \dot{\theta}_2$). These terms correspond to the Coriolis force, representing a lateral deflection force arising when the relative angle between the two swinging pendulums changes.
\end{itemize}

These ``forces'' are not external physical interactions but are purely geometric consequences of applying analytical mechanics within a rotating coordinate system. If the full non-linear equations of motion are expressed in the matrix form $M(\mathbf{\theta})\ddot{\mathbf{\theta}} + C(\mathbf{\theta}, \dot{\mathbf{\theta}})\dot{\mathbf{\theta}} + G(\mathbf{\theta}) = 0$, the matrix $C$ is explicitly recognized as the Coriolis and centrifugal force matrix.

Furthermore, the non-separable nature of the system has profound implications for numerical simulations. Because the kinetic energy is coordinate-dependent, standard explicit symplectic integrators (such as the semi-implicit Euler or velocity Verlet algorithms) cannot perfectly preserve symplecticity, inevitably leading to energy drift over long integration periods. To guarantee strict energy conservation and maintain high precision in such non-separable systems, fully implicit symplectic algorithms, such as the Gauss-Legendre Runge-Kutta methods, must be employed.

\subsection{Equations of Motion and Small-Angle Approximation}
The dynamics of the system are governed by the Euler-Lagrange equations:
\begin{align*}
\frac{d}{dt}\left(\frac{\partial L}{\partial \dot{\theta}_i}\right) - \frac{\partial L}{\partial \theta_i} &= 0, \quad \text{for } i = 1, 2.
\end{align*}

Evaluating these partial derivatives yields a set of highly non-linear, coupled second-order differential equations. For arbitrary angles, these equations exhibit chaotic behavior and can only be solved numerically. However, to understand the fundamental oscillatory properties, we analyze the system under the small-angle approximation, where $\theta_1, \theta_2 \ll 1$. Applying Taylor series expansions, we approximate $\sin\theta \approx \theta$ and $\cos\theta \approx 1 - \frac{\theta^2}{2}$, while ignoring higher-order terms such as $\dot{\theta}^2\theta$. The linearized equations of motion become:
\begin{equation}
\begin{aligned}
(m_1 + m_2)l_1 \ddot{\theta}_1 + m_2 l_2 \ddot{\theta}_2 + (m_1 + m_2)g \theta_1 &= 0, \\
m_2 l_1 \ddot{\theta}_1 + m_2 l_2 \ddot{\theta}_2 + m_2 g \theta_2 &= 0.
\end{aligned}
\end{equation}

\subsection{Derivation of Natural Frequencies}
The full double-pendulum equations are non-linear, so the angular frequencies discussed in this subsection are not exact global frequencies of the original system. Instead, they are obtained by first linearizing the non-linear equations under the small-angle approximation and then solving the resulting linear eigenvalue problem. In this linear regime, we seek oscillatory solutions of the form $\theta_1(t) = A_1 e^{i\omega t}$ and $\theta_2(t) = A_2 e^{i\omega t}$, where $\omega$ is the angular frequency and $A_1, A_2$ are the amplitudes. Substituting these into the linearized equations yields:
\begin{equation}
\begin{pmatrix}
(m_1 + m_2)g - (m_1 + m_2)l_1\omega^2 & -m_2 l_2 \omega^2 \\
-m_2 l_1 \omega^2 & m_2 g - m_2 l_2 \omega^2
\end{pmatrix}
\begin{pmatrix}
A_1 \\ A_2
\end{pmatrix} = \begin{pmatrix} 0 \\ 0 \end{pmatrix}.
\end{equation}

For non-trivial solutions, the determinant of the coefficient matrix must be zero. Writing $x=\omega^2$, the characteristic equation becomes
\begin{equation}
m_1 l_1 l_2 x^2 - (m_1 + m_2)g(l_1 + l_2)x + (m_1 + m_2)g^2 = 0.
\end{equation}
Hence, for the general case with arbitrary masses and rod lengths, the squared angular frequencies are
\begin{equation}
\omega_{\pm}^2 =
\frac{g\left[(m_1 + m_2)(l_1 + l_2) \pm \sqrt{(m_1 + m_2)\left((m_1 + m_2)(l_1 + l_2)^2 - 4m_1 l_1 l_2\right)}\right]}{2m_1 l_1 l_2}.
\end{equation}
These expressions are the generalized small-angle angular frequencies obtained from the linearized form of the originally non-linear system. For simplicity, if we consider the special case where $m_1 = m_2 = m$ and $l_1 = l_2 = l$, the frequencies simplify to:
\begin{equation}
\omega_{\pm}^2 = \frac{g}{l} \left( 2 \pm \sqrt{2} \right).
\end{equation}
These two frequencies, $\omega_+$ and $\omega_-$, represent the two fundamental modes of vibration for the double pendulum.

\subsection{Physical Implications of Normal Modes}
The normal modes represent specific patterns of motion where all parts of the system oscillate with the same frequency and a fixed phase relation. 

\begin{itemize}[leftmargin=*]
    \item \textbf{The In-Phase Mode (Lower Frequency, $\omega_-$):} Corresponding to $\omega_-^2 = (2 - \sqrt{2})\frac{g}{l}$, the amplitude ratio $A_2/A_1$ is positive (specifically, $\sqrt{2}$). Physically, this means both pendulum masses swing in the same direction simultaneously. The lower string remains relatively aligned with the upper string, reducing the overall restoring force from gravity. Consequently, this mode oscillates more slowly.
    \item \textbf{The Out-of-Phase Mode (Higher Frequency, $\omega_+$):} Corresponding to $\omega_+^2 = (2 + \sqrt{2})\frac{g}{l}$, the amplitude ratio $A_2/A_1$ is negative ($-\sqrt{2}$). In this mode, the two masses swing in opposite directions. This motion significantly bends the joint between the two rods, maximizing the restoring forces, which leads to a much faster, higher-frequency oscillation.
\end{itemize}
According to the superposition principle, any small-amplitude motion of the double pendulum can be described as a linear combination of these two fundamental normal modes. This theoretical foundation is crucial for verifying the accuracy of the numerical simulations presented in the following sections.

\subsection{Superposition State in the Linear Regime}
In the small-angle linear regime, the general solution is the superposition of the two normal modes:
\begin{equation}
\begin{aligned}
\theta_1(t) &= c_- A_{1,-}\cos(\omega_- t + \phi_-) + c_+ A_{1,+}\cos(\omega_+ t + \phi_+), \\
\theta_2(t) &= c_- A_{2,-}\cos(\omega_- t + \phi_-) + c_+ A_{2,+}\cos(\omega_+ t + \phi_+),
\end{aligned}
\end{equation}
where $c_{\pm}$ are mode amplitudes set by initial conditions, $\phi_{\pm}$ are initial phases, and $(A_{1,\pm}, A_{2,\pm})$ are eigenvectors associated with each normal mode. This form shows that the observed trajectory is not usually a pure in-phase or out-of-phase oscillation, but a mixed state containing both. When both coefficients are non-zero, energy is distributed across modes and the measured motion may show modulation and phase-mixing effects. The superposition description remains accurate only while linear assumptions hold; for larger amplitudes, non-linear coupling progressively invalidates pure modal superposition and leads to strongly irregular motion.

\subsection{Active Control and Rapid Amplitude Reduction}
In practical engineering applications, it is often desirable to suppress the oscillations of a double pendulum rapidly. Relying solely on passive natural damping (such as air drag and joint friction) is typically insufficient. By introducing an external control torque, we can actively extract energy from the system. Below are several theoretical approaches to achieve this in the small-angle regime.

\subsubsection{Active Damping via Velocity Feedback}
The most direct method is analogous to adding a viscous damper to a single pendulum. The linearized multi-degree-of-freedom equations of motion can be written in matrix form as:
\begin{equation}
M \ddot{\mathbf{\theta}} + C \dot{\mathbf{\theta}} + K \mathbf{\theta} = \mathbf{u}(t),
\end{equation}
where $M, C, K$ are the mass, damping, and stiffness matrices respectively, $\mathbf{\theta} = [\theta_1, \theta_2]^T$, and $\mathbf{u}$ is the control input. If we apply a control law $\mathbf{u} = -K_v \dot{\mathbf{\theta}}$ (where $K_v$ is a positive definite velocity gain matrix), we effectively increase the system's damping matrix to $C + K_v$. This active damping simultaneously dissipates energy from both normal modes, causing the vibrations to decay exponentially. While simple to implement, excessively high gains can lead to an overdamped, sluggish response.

\subsubsection{Anti-Phase Excitation and Dynamic Vibration Absorption}
Another approach draws inspiration from driven single pendulums. In a double pendulum, we can utilize one of the masses to suppress the vibration of the other. 
If an actuator is placed at the first pivot, it can apply a force perfectly out-of-phase with the measured velocity of the first pendulum, essentially acting as localized active damping.
A more elegant strategy is to use the second pendulum as an \emph{active dynamic vibration absorber} (similar to a Tuned Mass Damper, TMD). By controlling the torque at the second joint, the virtual natural frequency of the second pendulum can be tuned to match the current vibration frequency of the primary system. The vibrational energy of the first pendulum is rapidly transferred to the second pendulum, where it can be dissipated by the joint's damping.

\subsubsection{Optimal Control (LQR) and Modal Decoupling}
Because the double pendulum has two distinct normal modes (low-frequency in-phase and high-frequency out-of-phase), simple single-joint feedback might inadvertently suppress one mode while exciting the other. To address this modal coupling, modern control theory offers the Linear Quadratic Regulator (LQR).

In the Lagrangian framework, control action is introduced through the generalized forces $\mathbf{Q} = [Q_1, Q_2]^T$, so that the equations of motion become
\begin{equation}
\frac{d}{dt}\left(\frac{\partial L}{\partial \dot{\theta}_i}\right) - \frac{\partial L}{\partial \theta_i} = Q_i, \qquad i=1,2.
\end{equation}
After linearization about the equilibrium position and defining the state vector as $\mathbf{x} = [\theta_1, \theta_2, \dot{\theta}_1, \dot{\theta}_2]^T$, the controlled system can be written in state-space form as $\mathbf{\dot{x}} = A\mathbf{x} + B\mathbf{Q}$. The LQR method then seeks a state-feedback law of the form $\mathbf{Q} = -K\mathbf{x}$ that minimizes the cost function:
\begin{equation}
J = \int_{0}^{\infty} \left( \mathbf{x}^T Q \mathbf{x} + \mathbf{Q}^T R \mathbf{Q} \right) dt.
\end{equation}
By solving the Algebraic Riccati Equation, we obtain the optimal gain matrix $K$. This shifts the closed-loop poles deeper into the left half of the complex plane, ensuring the fastest possible decay of both modes simultaneously without exceeding the physical limits of the actuators. Interpreted physically, the controller computes the generalized forces required at the joints to remove energy from both modes in a coordinated way. If both joints are equipped with motors and sensors, LQR based on generalized-force feedback is one of the most reliable and rapidly convergent engineering solutions for amplitude reduction.

\subsection{Using a Simple Pendulum to Identify Air Drag: Theoretical Framework}

\textbf{Part I: Theoretical Framework of the Damped Pendulum}
\begin{enumerate}
    \item Under the small-angle approximation, the differential equation of motion for a damped simple pendulum is given by:
    \begin{equation}
    \ddot{\theta} + 2\beta\dot{\theta} + \omega_0^2\theta = 0
    \end{equation}
    \item Assuming a weak damping condition ($\beta < \omega_0$), the general solution to this equation is:
    \begin{equation}
    \theta(t) = A_0 \exp(-\beta t) \cos(\omega_d t + \phi_0)
    \end{equation}
\end{enumerate}

\textbf{Part II: Physical Significance of Key Parameters}
\begin{itemize}
    \item \textbf{Damping Coefficient ($\beta$)}: Measures the rate of energy dissipation from the system due to air drag and friction.
    \item \textbf{Undamped Natural Frequency ($\omega_0$)}: The inherent frequency at which the pendulum would oscillate if there were no damping ($\sqrt{g/l}$).
    \item \textbf{Time Constant ($\tau = 1/\beta$)}: The time required for the oscillation amplitude to decay to $1/e$ (approx. 37\%) of its initial value.
    \item \textbf{Damping Ratio ($\zeta = \beta/\omega_0$)}: A dimensionless parameter that indicates whether the system is underdamped, critically damped, or overdamped.
    \item \textbf{Relationship between Damped Period ($T_d$) and Undamped Period ($T_0$)}: $T_d = 2\pi / \sqrt{\omega_0^2 - \beta^2}$. This indicates that the presence of damping causes the actual oscillation period to become longer.
\end{itemize}

\textbf{Part III: Damping Measurement Method (0.37$\times$ Amplitude Decay)}
\begin{enumerate}
    \item According to the general solution, the amplitude envelope decays exponentially over time as:
    \begin{equation}
    A(t) = A_0 e^{-\beta t}
    \end{equation}
    \item When the amplitude decays to $1/e$ (approximately 0.37 times) of its initial value $A_0$, the elapsed time is exactly $t = \tau$ (or $1/\beta$).
\end{enumerate}

\textbf{Part IV: Experimental Procedure and Calculation}
\begin{itemize}
    \item \textbf{Step 1}: Release the pendulum from a small initial angle (less than 5 to 10 degrees) and record the initial maximum amplitude $A_0$.
    \item \textbf{Step 2}: Start the timer and observe the decay. Record the elapsed time $t_1$ when the maximum amplitude decreases to $0.37 \times A_0$.
    \item \textbf{Step 3}: Repeat the measurement multiple times to calculate the average decay time $\bar{t}$.
    \item \textbf{Step 4}: Using the measured average time constant $\tau = \bar{t}$, calculate the damping coefficient as $\beta = 1 / \bar{t}$.
\end{itemize}

\textbf{Part V: Experimental Precautions}
\begin{itemize}
    \item The theoretical derivation relies on the small-angle approximation.
    \item During the decay measurement, try to avoid interference such as external drafts, initial push, or non-planar (conical) swinging motion.
\end{itemize}

\section{Simulation}

\section{Experiment Results}
  Experiment results content...

  \section{Conclusion}
  Conclusion content...

% References (AQA Harvard Style)
\section{References}
\begin{thebibliography}{99}
  % Book Reference
  \bibitem{book1}
  Smith, J. (2020). \emph{Title of the Book}. Publisher Name.
  
  % Journal Article Reference
  \bibitem{journal1}
  Johnson, A. B. (2019). Title of the article. \emph{Journal Name}, 15(3), 45-67.
  
  % Website Reference
  \bibitem{web1}
  Organisation Name. (2021). Title of web page. Retrieved from https://www.example.com
  
  % Chapter in Edited Book
  \bibitem{chapter1}
  Brown, C. (2018). Title of chapter. In A. Editor \& B. Co-editor (Eds.), \emph{Title of Book} (pp. 123-145). Publisher Name.
  
  % Conference Paper
  \bibitem{conference1}
  Wilson, D. (2017). Title of paper. In \emph{Proceedings of the 10th Conference on Topic} (pp. 67-78). Conference Location.
\end{thebibliography}

% Appendices (Optional)
\appendix
\section{Appendix A}
  Appendix content...

\end{document}
